The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2019-06-19 razzo-methods.tex
Updates from Overleaf

\subsection{Model}
\begin{itemize}

\item 
\richel{TODO: move to Introduction}
Current phylogenetic tools assume that only a single speciation event 
can occur at any given time.
While this assumption is useful to construct a wide variety of successful 
models (e.g \cite{Maddison2007biSSE}, \cite{Valente2015}, 
\cite{etienne2012diversity}, \cite{etienne2014estimating}),
they disallow for environmental changes that trigger speciations 
in multiple clades at a same point in time.

\item 
\richel{TODO: move to Introduction}
In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates present 
also in the standard BD model. 
Additionally, MBD relies on two additional parameters. 
Parameter $\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, 
all species present in the phylogeny at that moment
have a probability $q$ to speciate at that time, which is 
independent on $\lambda$. 
Polytomies are not allowed in such process 
as each species can speciate only once at the time.

\item 
\richel{TODO: move to Introduction}
It is also possible to write down a likelihood function 
for such processes as in \cite{mbd}.

\end{itemize}

\subsection{Simulations}
\begin{itemize}

\item To investigate the effect of $s$ on $e$, we simulate phylogenies
for different values of $s$ spread equally from zero (no multiple-birth
event, thus a BD model) to one (species are created only from multiple-birth
events) and three intermediate values of 0.25, 0.5 and 0.75. Aggregating all
components of $s$, we show the effect of $s$ on $e$ in figure \ref{fig:1}.

\item To see the effect of the number of taxa on the relation 
between $s$ and $e$, we simulate phylogenies
for different number of extant taxa $n$. We use values of $n$
of 50, 100 and 200 extant taxa, as this will result in phylogenies
that are big enough to be useful, yet small of enough to
be computationally feasible. For the different values of $n$, we show
the relation between $s$ and $e$ in figure \ref{fig:5} and its effect
on the variance of $e$ in figure \ref{fig:6}.

\item To see the effect of the extinction rate on the relation 
between $s$ and $e$, we simulate phylogenies
with different extinction rates $\mu$. We use values of $\mu$
of 0.0, 0.1 and 0.2. An extinction rate of zero has two features:
(1) the model falls back to a pure multiple-birth model, 
(2) all multiple-birth events are observed.
For the different values of $\mu$, we show
the relation between $s$ and $e$ in figure \ref{fig:2}.

\item We start simulating $N_{S} = 1000$ MBD trees,
with either 50, 100 and 200 taxa.

\item From each MBD tree, a DNA sequence alignment is simulated. 
For each sequence alignment we then perform a Bayesian analysis 
to recover a posterior distribution of trees, 
each composed of $N_{P}$ phylogenies. 
Such analysis is performed using 
the 'pirouette' package (\cite{pirouette}) to call the BEAST2 tool 
suite from R. 
We let the Bayesian analysis assume a BD prior in both cases, 
to investigate the extent of the error we make under this assumption.

\item For each tree generated under the MBD model 
we aim to generate a "twin" tree under the BD model. 
With the word "twin" 
we denote a tree generated starting from the respective MBD tree, 
in order to perform a fair comparison with it. 
This operation has to be done, 
because we want to compare two trees 
that are generated by different processes. 
To do so we infer the parameters $\lambda_{BD}$ and $\mu_{BD}$ 
from the MBD maximizing the likelihood under a BD model. 
To perform this operation we use the function "\texttt{bd\_ML}" 
from the package "\texttt{DDD}" (\cite{etienne2012diversity}).

\item We then exploit such parameters to generate a BD tree 
using the function "\texttt{tess.sim.taxa.age}" 
from the package "\texttt{TESS}" (\cite{Hoehna2013}). 
We simulate the tree in such a way the new tree 
has the same number of tips and the same crown age as the MBD tree. 
We furthermore require that the BD tree conserve the topology of the MBD tree.
We have hypothesis H4 that, compared to the MBD trees, 
the error will be less in the BD twin tree.
The difference between the errors made in MBD and twin BD trees indicates
the impact the MBD process has on the error we make in inference using a
contemporary BD prior.

\include{razzo-fig_h_4.tex}

We want the MBD and twin BD trees to contain the same amount of information, 
i.e. the same number of DNA mutations and the same number of taxa at the present:

\begin{equation}
m_{MBD} = m_{BD} \label{m equivalence}
\end{equation}

The expected number of mutations $m$ of a phylogeny 
with crown age $-T$ (with $T>0$) in fact is given by
\richel{
  So one of use likes '-T', the other likes 'T'. How to resolve this?
}

\begin{equation}
m = L \cdot \rho \cdot \int_{0}^{T} n(t)\ dt \label{m calculation}
\end{equation}

where $L$ is the number of DNA nucleotides, 
$\rho$ is the per-site per-species mutation rate and
$n(t)$ the number of species at each time.

The parameter we'll tune is $\rho$ ... \richel{elaborate here :-)}

Since we cannot know $n_{BD}(t)$ before running simulations
we need to replace it with a proxy. 
For this reason we will use the average number of
species in time according to the BD model. 
It's well known that this is equal to \gio{insert proper citation}

\begin{equation}
    <n_{BD}>(t) = n_{0} \cdot e^{(\mu_{BD} - \lambda_{BD})t} \label{BD average n}
\end{equation}

where $n_{0} = n_{BD}(-T) = n_{MBD}(-T)$ is the initial number of species 
at the crown age.
From \ref{m equivalence}, \ref{m calculation} and \ref{BD average n} follows:

\begin{equation}
    m_{MBD} = 
    L \cdot \rho \cdot \int_{0}^{T} <n_{BD}>(t)\ dt 
    \\ =
    L \cdot \rho \cdot n_{0} \cdot \left[ \frac{e^{(\mu_{BD} - \lambda_{BD})T} - 1}{\mu_{BD} - \lambda_{BD}} \right]
\end{equation}

If we set $\mu_{BD} = \mu_{MBD}$ and reverse this relation 
we can extrapolate the value of $\lambda_{BD}$ to use to generate BD trees.

\input{razzo-fig_mbd.tex}

\item We explained how we set the parameters for each twin BD tree. 
Using this rules we generate a BD dataset. 
We repeat the analysis, producing alignments for each tree 
and subsequently using BEAST to produce a posterior for each of them.

\subsection{Measuring the inference error}

\item So far we have simulated two datasets of trees under the two models: 
$\{T_{i}^{BD}\}_{i=1}^{N_{S}}$ and $\{T_{i}^{MBD}\}_{i=1}^{N_{S}}$.
We used them to generate a dataset of alignments for each model: $\{X^{BD}_{i}\}_{i=1}^{N_{S}}$ and $\{X^{MBD}_{i}\}_{i=1}^{N_{S}}$. From each dataset we produced a posterior distribution from a BD prior: 
$P_{i}(\theta | X^{BD}_{i}, BD)$ and $P_{i}(\theta | X^{MBD}_{i}, BD)$.
\gio{
  1) We might want to rename the models, e.g. BD = (0) and MBD = (1). 
  These names with capital letters are too big and ugly;
}
\richel{
  I would strongly prefer MBD and BD, as I feel replacing the big ugly 
  capital letters by short pretty numbers hurts readability even more 
}

\item To compare the results for the two models we measure the inference 
error using the nLTT statistic between known/true tree and 
posterior/inferred trees (\cite{nltt}). 
To obtain such statistics the procedure is the following:

- From each tree $T_{i,j}^{M}$ (with $j=1,...,N_{S}$) 
  belonging to the posterior $P_{i}(\theta | X^{M}_{i}, BD)$ 
  and relative to the model $M$, we extrapolate the lineage-through-time (LTT), 
  in other words we measure the number of species as a function of 
  time $n_{i,j}(t)$. To allow a comparison we normalize dividing by the 
  maximum number of species of each tree, i.e. the number of tips at the 
  present $N_{i,j}(t)=\frac{n_{i,j}(t)}{n^{max}_{i,j}}$. We then define the 
  nLTT measure as

$nLTT_{i,j} = \int_{0}^{T} | N_{i,j}(t) - N_{T_{i}} | dt$

\gio{I am running out of letters :(}
\richel{Haha! I suggest to use the same equation and symbols 
  as equation 1 in
  the nLTT article of Janzen, Hoehna and Etienne, 2015:
}

$$
\Delta nLTT = \int_{0}^{1} | nLTT_1(t) - nLTT_2{t} | dt
$$

\subsection{Model selection}

@@

From these four marginal likelihood estimates, we calculate the weight of
the generative model and plot this in figure 2. We do this for both the 
alignments derived from the MBD tree and the BD twin tree. We expect that
the generative model has the heighest weight in both the MBD and BD alignments.
We expect this weight to be higher in the BD alignments.

\end{itemize}

>>>>

\subsection{Model}

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates present 
also in the standard BD model. 
Additionally, MBD relies on two additional parameters. 
Parameter $\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, each species present in the phylogeny at that moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature respect to the one triggered by $\lambda$. In fact, whereas parameter $\lambda$ can be seen as describing a sympatric process, $\nu$ induces the rise of geographical barriers interrupting the gene flow and leading to an allopatric speciation.
Even though multiple speciations can co-occur, polytomies are not allowed in such process as each species can speciate only once at the time.
A likelihood expression for the process is provided in \cite{mbd}.

\subsection{Tree simulations}

We can easily simulate such processes in continuous time using the Doob-Gillespie algorithm. Simulations are performed using the function "mbd\_sim" from the \verb;mbd; package (\citep{mbd}).
We let parameters vary using all possible combinations of value included in Table~\ref{tab:simulation_parameters} to simulate $1000$ independent trees for each parameter setting.

\input{razzo-tables/tab_simparameters.tex}

\subsection{Pirouette}

Once the MBD dataset has been simulated we exploit the \verb;pirouette; package (\citep{pirouette}) to assess the error made by BEAST2, executing  the inference using a BD tree prior.

We briefly summarize here how the \verb;pirouette; routine works.
From each MBD tree a DNA sequence alignment is simulated. 
For each sequence alignment we then perform a Bayesian inference analysis (using BEAST2) to recover a posterior distribution of trees. 
For each parameter setting, this process leads to an inference error distribution. To evaluate the extent of this error we also run a \verb;pirouette; "twin" pipeline, through which we mimic the original pipeline in every aspect but starting instead from "twin" trees.
From each MBD simulated tree we can produce its BD twin, inferring the most likely BD parameters (i.e. $\lambda$ and $\mu$) through maximum likelihood, and use them to simulate a standard BD tree (see \cite{pirouette} for more information), keeping the topology of the original tree.
We let the Bayesian analysis assume a BD prior in both cases, to investigate the extent of the error we make under this assumption. The twin pipeline serves as an estimation of the baseline error, as its error it is not due by the mismatch of the generative prior with the inference prior.

\iffalse
\subsection{Model selection}

@@

From these four marginal likelihood estimates, we calculate the weight of
the generative model and plot this in figure 2. We do this for both the 
alignments derived from the MBD tree and the BD twin tree. We expect that
the generative model has the heighest weight in both the MBD and BD alignments.
We expect this weight to be higher in the BD alignments.
\fi
